---
title: "Causal Mediation"
author: |
  | Mark Andrews
  | Psychology Department, Nottingham Trent University
  | 
  | \faEnvelopeO\  ```mark.andrews@ntu.ac.uk```
fontsize: 10pt
output:
 beamer_presentation:
  keep_tex: true
  fonttheme: "serif"
  includes:
   in_header: preamble.tex
---

```{r, echo=F}
knitr::opts_chunk$set(echo = F, prompt = F, warning = F, message = F, comment='#>')
# Thanks to 
# https://github.com/ramnathv/slidify/issues/189#issuecomment-15850008
hook1 <- function(x){ gsub("```\n*```r*\n*", "", x) }
hook2 <- function(x){ gsub("```\n+```\n", "", x) }
knitr::knit_hooks$set(document = hook1)
```

```{r}
library(tidyverse)
library(magrittr)
theme_set(theme_classic())
```


# Establishing causal effects

* We hypothesize that $x$ causes $y$, and we aim to establish its causal effect. 

\begin{center}
\begin{tikzpicture}
\tikzstyle{every path} = [-latex,thick]
\tikzstyle{every node} = [rectangle, fill=gray!30]
\node (x) at (0, 0) {$x$};
\node (y) at (2, 0) {$y$};
\draw (x) -- (y);
\draw (x) -- (y);
\end{tikzpicture}
\end{center}

* For example, $y$ could be getting lung cancer, $x$ could be smoking 20 cigarettes a day every day from an early age.

* For person $i$, $y_i \in \{0, 1\}$ is whether they get lung cancer and $x_i \in \{0, 1\}$ is whether they are a smoker (as just defined). 

# Confounds

* Of course, a third variable (or set of variables) $z$ may also affect the value of $x$ and $y$.

\begin{center}
\begin{tikzpicture}
\tikzstyle{every path} = [-latex,thick]
\tikzstyle{every node} = [rectangle, fill=gray!30]
\node (x) at (0, 0) {$x$};
\node (y) at (2, 0) {$y$};
\node[fill=white] (z) at (1, 1) {$z$};
\draw (x) -- (y);
\draw (z) -- (x);
\draw (z) -- (y);
\end{tikzpicture}
\end{center}

* For example, $z \in (-\infty, \infty)$ could be bad attitude towards healthy lifestyles generally. For example, if $z$ is high, the person may have poor diet, lack of exercise, drink heavily, and smoke heavily.

* We will assume that we do not know $z$.

# Potential outcomes framework

* In the *potential outcomes framework*, we consider what would have happened to any person $i$ had they been *assigned* $x_i = 0$ or $x_i = 1$.

* If $x_i$ was assigned the value of $0$, then we say the potential outcomes of $y_i$ is $y_i(0)$. Likewise, if $x_i$ was assigned the value of $1$, then we say the potential outcomes of $y_i$ is $y_i(1)$. 

* For person $i$, they always have two potential outcomes, $(y_i(0), y_i(1))$ and the causal effect of $x_i$ is 
$$
\delta_i = y_i(1) - y_i(0).
$$

* However, for any person $i$ we observe 
$$
y_i = 
\begin{cases} 
  y_i(0)\quad\text{if $x_i = 0$},\\
  y_i(1)\quad\text{if $x_i = 1$}.
\end{cases}
$$

* We only ever observe either $y_i(1)$ or $y_i(0)$ and never both, and so can never know $\delta_i$. This is known as the *fundamental problem of causal inference*.

# The power of randomization

* In observational studies, $(y_i(0), y_i(1))$ is not statistically independent of $x_i$.
* However, if $x_i$ is assigned randomly, then $(y_i(0), y_i(1))$ *is* statistically independent of $x_i$.
* We can write this as follows:
$$
(y_i(0), y_i(1)) \CI x_i
$$
which means
$$
\begin{aligned}
\Prob{y_i(1) \given x_i} &= \Prob{y_i(1)} ,\\
\Prob{y_i(0) \given x_i} &= \Prob{y_i(0)} .
\end{aligned}
$$

# Average treatment effect (ATE)

* The *average treatment effect* (ATE) is the difference of average of the two potential outcomes.
$$
\begin{aligned}
\langle \delta \rangle &= \lim_{n \to \infty} \frac{1}{n} \sum_{i=1}^n \delta_i,\\
                       &= \lim_{n \to \infty} \frac{1}{n} \sum_{i=1}^n \left(y_i(1) - y_i(0)\right),\\
                       &=  \left(\lim_{n \to \infty}\frac{1}{n} \sum_{i=1}^n y_i(1)\right)
                         - \left(\lim_{n \to \infty}\frac{1}{n} \sum_{i=1}^n y_i(0)\right),\\
                       &= \left\langle y(1) \right\rangle - \langle y(0) \rangle
\end{aligned}
$$

# Average treatment effect under randomization

* When $x_i$ is randomly assigned
$$
\langle y(0) \rangle = 
  \lim_{n \to \infty} \frac{1}{n} \sum_{i=1}^n \Prob{y_i \given x_i = 0} =
  \hat{y}_{[x_i = 0]}
$$
and 
$$
\langle y(1) \rangle = 
  \lim_{n \to \infty} \frac{1}{n} \sum_{i=1}^n \Prob{y_i \given x_i = 1} =
  \hat{y}_{[x_i = 1]}
$$

* Therefore,
$$
\left\langle y(1) \right\rangle - \langle y(0) \rangle =
\hat{y}_{[x_i = 1]} - \hat{y}_{[x_i = 0]}
$$

# Mediation

* In a mediation model, we hypothesize that the effect of $x$ on $y$ is, at least partially, mediated by $m$.

\begin{center}
\begin{tikzpicture}
\tikzstyle{every path} = [-latex,thick]
\tikzstyle{every node} = [rectangle, fill=gray!30]
\node (x) at (0, 0) {$x$};
\node (y) at (2, 0) {$y$};
\node (m) at (1, 1) {$m$};
\draw (x) -- (y);
\draw (x) -- (m);
\draw (m) -- (y);
\end{tikzpicture}
\end{center}

* For example, to use a new example, attending a university ($x$) leads to high income ($y$) partially because it increases a person's social network ($m$).

# Mediation and potential outcomes

* Assuming that $x_i$ is binary, the potential values of the mediator are $(m_i(0), m_i(1))$. 
* Thus, the potential outcomes are
$$
y_i(0, m_i(0)), 
y_i(0, m_i(1)),
y_i(1, m_i(0)),
y_i(1, m_i(1)).
$$

* For example, $y_i(1, m_i(0))$ is the value that $y_i$ would take if $x_i = 1$ and $m_i$ had the value it would take if $x_i = 0$. 

* Using the concrete example,  $y_i(1, m_i(0))$ is the income that person $i$ would obtain had they been assigned to attend university, i.e. $x_i = 1$, but their social network is that which would have obtained had they not attended university.

# Causal mediation, or indirect, effect

* The effect of the mediator when $x_i = 0$ is
$$
\delta_i(0) = y_i(0, m_i(1)) - y_i(0, m_i(0))
$$ 
which is the difference in the outcome that person $i$ would have obtained had they not gone to university and $m_i$ took the value of $m_i(1)$ rather than $m_i(0)$.

* The effect of the mediator when $x_i = 1$ is
$$
\delta_i(1) = y_i(1, m_i(1)) - y_i(1, m_i(0))
$$
which is the difference in the outcome that person $i$ would have obtained had they gone to university and $m_i$ took the value of $m_i(1)$ rather than $m_i(0)$.

* The *no-interaction* assumption is that 
$$\delta_i = \delta_i(0) = \delta_i(1).$$


# The direct effect

* The effect of the treatment when $m_i = m_i(0)$ is
$$
\zeta_i(0) = y_i(1, m_i(0)) - y_i(0, m_i(0))
$$ 
which is the difference in the outcome that person $i$ would have obtained had they gone to university compared to if they had not, assuming $m_i$ took the value of $m_i(0)$.

* The effect of the treatment when $m_i = m_i(1)$ is
$$
\zeta_i(1) = y_i(1, m_i(1)) - y_i(0, m_i(1))
$$ 
which is the difference in the outcome that person $i$ would have obtained had they gone to university compared to if they had not, assuming $m_i$ took the value of $m_i(1)$.

* Again, the *no-interaction* assumption is that 
$$\zeta_i = \zeta_i(0) = \zeta_i(1).$$

# Confounders

* In general, there may be observed confounders $z$ that affect $x$, $m$, and $y$.
 
\begin{center}
\begin{tikzpicture}
\tikzstyle{every path} = [-latex,thick]
\tikzstyle{every node} = [rectangle, fill=gray!30]
\node (x) at (0, 0) {$x$};
\node (y) at (2, 0) {$y$};
\node (m) at (1, 1) {$m$};
\node (z) at (1, 2) {$z$};
\draw (x) -- (y);
\draw (x) -- (m);
\draw (m) -- (y);
\draw (z) to [bend right=30] (x);
\draw (z) to [bend left=30] (y);
\draw (z) -- (m);
\end{tikzpicture}
\end{center}

* In our example, family background or intelligence etc might be examples of such confounders.

# Sequential ignorability

* Sequential ignorability assumes:

  1. Potential outcomes and potential mediators are independent of treatment, conditional on covariates.
  1. Potential outcomes are independent of the mediators, conditional on the treatment.

* The first part holds if the treatment $x$ is randomized.
* The second part does not necessarily even when $x$ is randomized because the mediator is not randomized.

# A general estimating algorithm

* Under sequential ignorability, we may estimate the average causal mediation effects (ACME) and average direct effect (ADE).

* Fit separate models outcome and mediator:

  1. Outcome model: $\Prob{y_i \given x_i, m_i, z_i}$.
  1. Mediator model: $\Prob{m_i \given x_i, z_i}$.
  
  where $z_i$ are covariates.
  
* There is no restriction, e.g. no linearity, normality, etc, restrictions, on the nature of these regression models.

# A general estimating algorithm

* Using the mediator model, predict $m_i(0)$ and $m_i(1)$ as follows:

  1. $m_i(0) = \Prob{m_i \given x_i = 0, z_i}$
  1. $m_i(1) = \Prob{m_i \given x_i = 1, z_i}$
  
* Using the outcome model, predict $y_i(1, m_i(0))$ and $y_i(1, m_i(1))$ as follows:

  1. $y_i(1, m_i(0)) = \Prob{y_i \given x_i = 1, m_i = m_i(0), z_i}$
  1. $y_i(1, m_i(1)) = \Prob{y_i \given x_i = 1, m_i = m_i(1), z_i}$
  
* Calculate 
  $\hat{\delta}_i = y_i(1, m_i(1)) - y_i(1, m_i(0))$
  and then 
  $$
  \hat{\delta} = \frac{1}{n} \sum_{i = 1}^n \delta_i
  $$
  is our estimator of $\delta$, the average causal mediation effect (ACME).
  

# A general estimating algorithm

* Using the outcome model, predict $y_i(1, m_i)$ and $y_i(0, m_i)$ as follows:

  1. $y_i(1, m_i) = \Prob{y_i \given x_i = 1, m_i, z_i}$
  1. $y_i(0, m_i) = \Prob{y_i \given x_i = 0, m_i, z_i}$
    
* Calculate 
  $\hat{\zeta}_i = y_i(1, m_i) - y_i(0, m_i)$
  and then 
  $$
  \hat{\zeta} = \frac{1}{n} \sum_{i = 1}^n \zeta_i
  $$
  is our estimator of $\zeta$, the average direct effect (ADE).