---
title: "Causal models"
author: |
  | Mark Andrews
  | Psychology Department, Nottingham Trent University
  | 
  | \faEnvelopeO\  ```mark.andrews@ntu.ac.uk```
fontsize: 10pt
output:
 beamer_presentation:
  keep_tex: true
  fonttheme: "serif"
  includes:
   in_header: preamble.tex
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(magrittr)
library(knitr)
library(pander)

theme_set(theme_classic())
set.seed(101010)

```

# Causal \xdag

* A causal \xdag is a \xdag: A graph with vertices representing variables, and directed edges between the vertices and contain no cycles.
* It represents the causal relationships between variables: If one variable $x$ causes another $y$, we write $x \to y$.
* The absence of an direct edge between $x$ and $y$ indicates no *direct* effect of $x$ on $y$.
* The absence of directed edg between variables is as or more informative than the presence of directed edge.
* A causal \xdag must have all relevant variables, all explanatory and outcome variables and *all their common causes*.

# Causal \xdag

* A causal \xdag contains three key structures:

  * Chains
  * Forks
  * Colliders
  
# Chains

* Chains encode causal chains.

  \begin{tikzpicture}
  \tikzstyle{every path} = [-latex,thick]
  \tikzstyle{every node} = [circle, fill=gray!30]
  \node (x) at (0, 0) {$x$};
  \node (z) at (1, 1) {$z$};
  \node (y) at (2, 0) {$y$};
  \draw (x) -- (z);
  \draw (z) -- (y); 
  \end{tikzpicture}
  
* Controlling for or blocking $z$ blocks the causal effect of $x$ on $y$.

# Forks

* Forks indicate confounders or spurious correlations.
  
  \begin{tikzpicture}
  \tikzstyle{every path} = [-latex,thick]
  \tikzstyle{every node} = [circle, fill=gray!30]
  \node (x) at (0, 0) {$x$};
  \node (y) at (2, 0) {$y$};
  \node (z) at (1, 1) {$z$};
  \draw (z) -- (x);
  \draw (z) -- (y); 
  \end{tikzpicture}
  
* Controlling for or blocking $z$ deconfounds the spurious correlation between $x$ and $y$.
  
# Colliders
    
* Colliders encode *endogenous associations*.

  \begin{tikzpicture}
  \tikzstyle{every path} = [-latex,thick]
  \tikzstyle{every node} = [circle, fill=gray!30]
  \node (x) at (0, 1) {$x$};
  \node (z) at (1, 0) {$z$};
  \node (y) at (2, 1) {$y$};
  \draw (x) -- (z);
  \draw (y) -- (z); 
  \end{tikzpicture}
  
* Controlling for or blocking $z$ *creates* a spurious correlation between $x$ and $y$.

# d-separation

* Two variables $x$ and $y$ in a causal \xdag are d-separated by $z$ if they are conditionally independent, conditional on $z$. 
* We write this as follows:
$$
x \CI y \given z
$$
* If two variables are conditionally independent, they will have a partial correlation of zero.
* In general, we can have sets of variables being d-separated by set of other variables.
* Every causal \xdag leads to list of d-separations. 
* Each d-separation is a testable hypothesis.
* The minimal set of d-separations (i.e. from which all others are derived) is the *basis set*.

# d-separation: Alternative definitions

* Two variables $x$ and $y$ are d-separated given $z$ iff *each* path from $x$ to $y$ is d-separated by $z$.

* A path is d-separated by $z$ iff any of the following hold.

  * The path contains a chain $x \rightarrow z \rightarrow y$ or $x \leftarrow z \leftarrow y$.
  * The path contains a fork $x \leftarrow z \rightarrow y$.
  * The path contains a collider $x \rightarrow m \leftarrow y$, where $m$ is neither $z$ nor any of $m$'s dependents is $z$.

# d-separation: Alternative definitions

* The two variables $x$ and $y$ are d-connected if there is *any* active path between them, and are d-separated if *all* paths that connect them are inactive.

* A path between $x$ and $y$ is *blocked* by variable $z$ if any of the following hold:

  1. $x$ and $y$ are connected by a chain in which $z$ is a middle node.
  1. $x$ and $y$ are connected by a common cause, and $z$ is that common cause.
  1. $x$ and $y$ are connected by a common effect (*collider*), but $z$ is not that common effect, and $z$ is not one of the effects of the common effect.

# Deciding d-separation: Bayes ball

* We can decide if $x$ and $y$ are d-separated by $z$ as follows:

* Shade $z$ (or all nodes in $z$ if $z$ is a set).
* Place a "ball" in $x$, let it "bounce around" and see if it reaches $y$.
  1. The ball can pass through chain unless middle node is shaded.
  2. The ball can pass through a fork unless shaded in middle.
  3. The ball can *not* pass through collider unless middle node is shaded.
  
* If the ball can get from $x$ to $y$ then $x$ and $y$ are d-connected, otherwise they are d-separated.

# Deciding d-separation: The "moral" and disoriented graph

* Draw the ancestral graph consisting only of the relevant variables (e.g., $x$, $y$, $z$) and all of their *ancestors* (parents, parents’ parents, etc.)

* *Moralize* the ancestral graph by *marrying* the parents. For each pair of variables with a common child, draw an *undirected* edge between them. If a variable has more than two parents, draw lines between every pair of parents.
* *Disorient* the graph by replacing the directed edges with undirected edges.
* Delete the *givens* and their edges, i.e. erase the conditioning variables from the graph and erase all of their connections.
5. If the $x$ and $y$ variables are disconnected in this graph (no path between them), they are d-separated.

# Example 

\begin{center}
  \begin{tikzpicture}
  \tikzstyle{every path} = [-latex,thick]
  \tikzstyle{every node} = [circle, fill=gray!30]
    \node (a) at (1, 5) {$A$};
    \node (b) at (3, 5) {$B$};
    \node (c) at (2, 4) {$C$};
    \node (d) at (1, 3) {$D$};
    \node (e) at (3, 3) {$E$};
    \node (f) at (2, 2) {$F$};
    \node (g) at (1, 1) {$G$};
  \draw (a) -- (c);
  \draw (b) -- (c); 
  \draw (c) -- (d); 
  \draw (c) -- (e); 
  \draw (d) -- (f);
  \draw (f) -- (g);
  \end{tikzpicture}
\end{center}

# d-separation queries

1. $A \CI B \given D, F$ ?
1. $A \CI B$ ?
1. $A \CI B \given C$ ?
1. $D \CI E \given C$ ?
1. $D \CI E$ ?
1. $D \CI E \given A, B$ ?
1. $D \CI E \given C$?
1. $D \CI G \given C$?


# d-separation queries

1. $A \CI B \given D, F$? No
1. $A \CI B$? Yes
1. $A \CI B \given C$? No
1. $D \CI E \given C$? Yes
1. $D \CI E$? No
1. $D \CI E \given A, B$ No
1. $D \CI E \given C$? Yes
1. $D \CI G \given C$? No

# Using `bnlearn`

```{r, echo=T, message=F}
library(bnlearn)
bnet <- model2network("[a][b|f][c|e][e|a:b][f]")

plot(bnet)


```


# Using `bnlearn`

```{r, echo=T}
# Is a d-separated from b given c?
dsep(bnet, "a", "b", "c")

# Is a d-sepated from b given f
dsep(bnet, 'a', 'b', 'f')

```

# Using `bnlearn`
```{r, echo=T}
bnet <- model2network("[a][b][c|a:b][d|c][e|c][f|d][g|f]")

plot(bnet)
```


# Using `bnlearn`
```{r, echo=T}

# a d-separated from e given {}
dsep(bnet, 'd', 'e')

# a d-separated from e given {a,b}
dsep(bnet, 'd', 'e', c('a', 'b'))

```





# Using `bnlearn`
```{r, echo=T}

# d d-separated from e given c
dsep(bnet, 'd', 'e', 'c')

# a d-separated from e given {}
dsep(bnet, 'd', 'e')

# a d-separated from e given {a,b}
dsep(bnet, 'd', 'e', c('a', 'b'))

```

# Using `daggity`

```{r, echo=T, message =F}
library(dagitty)
g1 <- dagitty( "dag {
    Y <- A
    Y <- C
    Y <- F <- T <- B
    Y <- E <- D <- B
    Y <- D <- B
    C <- A
    C <- B
}")
```

# Using `daggity`


```{r, echo=T}
plot(graphLayout(g1))
```

# Using `dagitty`

* The *basis set* of d-separations.

```{r, echo=T}
impliedConditionalIndependencies(g1, type = 'basis.set')
```

