---
title: "Structural Equation Models"
author: |
  | Mark Andrews
  | Psychology Department, Nottingham Trent University
  | 
  | \faEnvelopeO\  ```mark.andrews@ntu.ac.uk```
fontsize: 10pt
output:
 beamer_presentation:
  keep_tex: true
  fonttheme: "serif"
  includes:
   in_header: preamble.tex
---

```{r, echo=F}
knitr::opts_chunk$set(echo = F, prompt = F, warning = F, message = F, comment='#>')
# Thanks to 
# https://github.com/ramnathv/slidify/issues/189#issuecomment-15850008
hook1 <- function(x){ gsub("```\n*```r*\n*", "", x) }
hook2 <- function(x){ gsub("```\n+```\n", "", x) }
knitr::knit_hooks$set(document = hook1)
```

```{r}
library(tidyverse)
library(magrittr)
library(modelr)
theme_set(theme_classic())
```

# What are Structural Equation Models?

* Structural equation models (\sem) are 

  1. Systems of statistical regression models \ldots
  1. \ldots that represent hypothetical causal models.
  
* As regression models, they generally include *latent* variables.

* Path analysis and \sem can be seen as either identical, or \sem can be seen as path analysis with additional latent variables.

# A very brief history

* \sem began with the path analysis work of biologist Sewall Wright around 1920.
* \sem became more widely used from 1970s onwards, but the focus became centred on algorithms and computer implementation, the peculiarities or linear-normal models, and the causal interpretation of \sem was downplayed or eschewed. 
* Since 2000s, \sem has become increasingly associated with *causal graph theory* (see J. Pearl), and no longer focused on linear-normal models.

# \sem & \xdag 

* A \sem model can be represented by a *directed acyclic graph* (\xdag), where vertices represent variables and directed edges represent causal relationships

* For example, 

\begin{center}
\begin{tikzpicture}
\tikzstyle{every path} = [-latex,thick]
\tikzstyle{every node} = [rectangle, fill=gray!30]
\node (x) at (0, 0) {$x$};
\node (y) at (2, 0) {$y$};
\node (m) at (1, 1) {$m$};
\node (z) at (1, 2) {$z$};
\draw (x) -- (y);
\draw (x) -- (m);
\draw (m) -- (y);
\draw (z) to [bend right=30] (x);
\draw (z) to [bend left=30] (y);
\draw (z) -- (m);
\end{tikzpicture}
\end{center}



* For example, we interpret the following to mean that $x$ causes $y$:

\begin{center}
\begin{tikzpicture}
\tikzstyle{every path} = [-latex,thick]
\tikzstyle{every node} = [rectangle, fill=gray!30]
\node (x) at (0, 0) {$x$};
\node (y) at (1, 0) {$y$};
\draw (x) -- (y);
\end{tikzpicture}
\end{center}

# Regression models and \textsc{dag}s

* Any regression model, or probabilistic model, can be represented by a \xdag.
* All variables, including parameters, are represented by vertices and directed edges represent conditional statistical dependence.
* A sem is often represented by a *nonparametric* \xdag, i.e., where the parameters and functional form of the regression model are not represented.
* Crucially, directed edges represent causal relationships.

# Overview of course

* Regression models
* Mediation models, including causal mediation
* Latent variable models
* \sem with `lavaan` and `blavaan`
* \sem with `stan`, `brms`, etc.

