---
title: "Mediation Analysis"
author: |
  | Mark Andrews
  | Psychology Department, Nottingham Trent University
  | 
  | \faEnvelopeO\  ```mark.andrews@ntu.ac.uk```
fontsize: 10pt
output:
 beamer_presentation:
  keep_tex: true
  fonttheme: "serif"
  includes:
   in_header: preamble.tex
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(magrittr)
library(knitr)
library(pander)

theme_set(theme_classic())
set.seed(101010)

```

# Mediation

* In a mediation model, the effect of one variable $x$ on another $y$ is due to its effect on a third variable $m$, which then affects $y$.

* Changes in the variable $x$ lead to changes in $m$ that then lead to changes in $y$.

* As an example of a mediation effect, it is widely appreciated that tobacco smoking $x$ raises the probability of lung cancer $y$, and that this effect is due to tar (tobacco residue) produced by the burning of the tobacco accumulating the lungs $m$. 

# Pure mediation

* In a *pure* or *full* mediation model, the effect of $x$ on $y$ is entirely due to its effect on $m$.

\begin{center}
\begin{tikzpicture}
\tikzstyle{every path} = [-latex,thick]
\tikzstyle{every node} = [circle, fill=gray!30]
\node (x) at (0, 0) {$x$};
\node (y) at (4, 0) {$y$};
\node (m) at (2,0) {$m$};
\draw (x) -- (m);
\draw (m) -- (y); 
\end{tikzpicture}
\end{center}

* Assuming that we are dealing with a normal linear model, we can write a pure mediation model as follows:
\begin{align*}
\text{for $i \ldots 1\ldots n$,}\quad y_i &\sim N(\mu^y_i, \sigma_y^2),
\quad \mu^y_i = \beta_{y0} + \beta_{ym} m_i,\\
m_i &\sim N(\mu^m_i, \sigma_m^2),
\quad \mu^m_i = \beta_{m0} + \beta_{mx} x_i,
\end{align*}


# Partial mediation

* In *partial mediation*, we assume that $x$ affects $m$ and $m$ affects $y$ as before, but there is also a direct effect of $x$ on $y$, as in the following diagram.

\begin{center}
\begin{tikzpicture}
\tikzstyle{every path} = [-latex,thick]
\tikzstyle{every node} = [circle, fill=gray!30]
\node (x) at (0, 0) {$x$};
\node (y) at (4, 0) {$y$};
\node (m) at (2, 2) {$m$};
\draw (x) -- (m);
\draw (m) -- (y); 
\draw (x) -- (y); 
\end{tikzpicture}
\end{center}

* The partial mediation model can be written as follows.
\begin{align*}
\text{for $i \ldots 1\ldots n$,}\quad y_i &\sim N(\mu^y_i, \sigma_y^2),
\quad \mu^y_i = \beta_{y0} + \beta_{ym} m_i + \beta_{yx} x_i,\\
m_i &\sim N(\mu^m_i, \sigma_m^2),
\quad \mu^m_i = \beta_{m0} + \beta_{mx} x_i,
\end{align*}

# Example 1

* In order to explore mediation models, let us begin with data generated according to a specific model.
```{r, echo=T}
N <- 100

b_m0 <- 1.25; b_mx <- 1.25; 
b_y0 <- -0.5; b_ym <- 1.75; b_yx <- 0.75;
sigma_m <- 1.5; sigma_y <- 2.0

mediation_df <- tibble(
  x = rnorm(N, sd = 2),
  m = b_m0 + b_mx * x + rnorm(N, sd = sigma_m),
  y = b_y0 + b_ym * m + b_yx * x + rnorm(N, sd = sigma_y)
)
```

# Using `lavaan`

* Let us now set up this model using `lavaan`.
```{r, echo=T, message=F}
library(lavaan)

mediation_model_spec_1 <- '
y ~ m + x
m ~ x
'
```
* For example, by writing `y ~ m`, we are assuming that for each $i$, $y_i = \beta_{y0} + \beta_{ym} m_i + \epsilon^y_i$.
* However, by default, unless we explicitly state in the formula that we are using an intercept term, as follows, we will not get information about it. 
```{r, echo=T}
mediation_model_spec_1 <- '
y ~ 1 + m + x
m ~ 1 + x
'
```

# Using `lavaan`

* Now we call `lavaan::sem` with reference to `mediation_model_spec_1`, and this fits the model using maximum likelihood estimation.
```{r, echo=T}
mediation_model_1 <- sem(mediation_model_spec_1, 
                         data = mediation_df)

parameterEstimates(mediation_model_1)
```

# Model comparison 

* In mediation analysis, a major aim is evaluating first whether there is evidence of a mediation of the effect of $x$ on $y$ by $m$, and then whether this is pure or partial mediation. 
* To do so, we first specify and then fit the full mediation model`.
```{r, echo=T}
mediation_model_spec_0 <- '
  y ~ 1 + m 
  m ~ 1 + x
'
mediation_model_0 <- sem(mediation_model_spec_0, 
                         data = mediation_df)

mediation_models <- c(model_0 = mediation_model_0,
                      model_1 = mediation_model_1)

map_dbl(mediation_models, AIC)
```
```{r}
stopifnot(AIC(mediation_model_1) < AIC(mediation_model_0))
```

# Alternative models: 1 to 4

\begin{figure}  
\centering  

\subfigure[]  
{  
\begin{tikzpicture}[show background rectangle]
\tikzstyle{every node} = [circle, fill=gray!30]
\tikzstyle{every path} = [-latex,thick]
\node (x) at (0, 0) {$x$};
\node (y) at (4, 0) {$y$};
\node (m) at (2, 2) {$m$};
\end{tikzpicture}
}  
\subfigure[]  
{  
\begin{tikzpicture}[show background rectangle]
\tikzstyle{every node} = [circle, fill=gray!30]
\tikzstyle{every path} = [-latex,thick]
\node (x) at (0, 0) {$x$};
\node (y) at (4, 0) {$y$};
\node (m) at (2, 2) {$m$};
\draw (x) -- (m);
\end{tikzpicture}
}
\subfigure[]  
{  
\begin{tikzpicture}[show background rectangle]
\tikzstyle{every node} = [circle, fill=gray!30]
\tikzstyle{every path} = [-latex,thick]
\node (x) at (0, 0) {$x$};
\node (y) at (4, 0) {$y$};
\node (m) at (2, 2) {$m$};
\draw (m) -- (y); 
\end{tikzpicture}
}
\subfigure[]  
{  
\begin{tikzpicture}[show background rectangle]
\tikzstyle{every node} = [circle, fill=gray!30]
\tikzstyle{every path} = [-latex,thick]
\node (x) at (0, 0) {$x$};
\node (y) at (4, 0) {$y$};
\node (m) at (2, 2) {$m$};
\draw (x) -- (y); 
\end{tikzpicture}
}

\end{figure}

# Alternative models: 5 to 8


\begin{figure}  
\centering  

\subfigure[]  
{  
\begin{tikzpicture}[show background rectangle]
\tikzstyle{every node} = [circle, fill=gray!30]
\tikzstyle{every path} = [-latex,thick]
\node (x) at (0, 0) {$x$};
\node (y) at (4, 0) {$y$};
\node (m) at (2, 2) {$m$};
\draw (x) -- (m);
\draw (m) -- (y); 
\end{tikzpicture}
}
\subfigure[]  
{  
\begin{tikzpicture}[show background rectangle]
\tikzstyle{every node} = [circle, fill=gray!30]
\tikzstyle{every path} = [-latex,thick]
\node (x) at (0, 0) {$x$};
\node (y) at (4, 0) {$y$};
\node (m) at (2, 2) {$m$};
\draw (x) -- (m);
\draw (x) -- (y); 
\end{tikzpicture}
}
\subfigure[]  
{  
\begin{tikzpicture}[show background rectangle]
\tikzstyle{every node} = [circle, fill=gray!30]
\tikzstyle{every path} = [-latex,thick]
\node (x) at (0, 0) {$x$};
\node (y) at (4, 0) {$y$};
\node (m) at (2, 2) {$m$};
\draw (m) -- (y);
\draw (x) -- (y); 
\end{tikzpicture}
}
\subfigure[]  
{  
\begin{tikzpicture}[show background rectangle]
\tikzstyle{every node} = [circle, fill=gray!30]
\tikzstyle{every path} = [-latex,thick]
\node (x) at (0, 0) {$x$};
\node (y) at (4, 0) {$y$};
\node (m) at (2, 2) {$m$};
\draw (x) -- (m);
\draw (m) -- (y); 
\draw (x) -- (y); 
\end{tikzpicture}
}
\end{figure}

# Model evalation

*  Read alternative models specifications from file:

```{r, echo=T}
source('mediation_models_specs.R')
```

* Having each model specified as an element of a list, we now use `purrr::map` to fit each model and calculate its \aic score.
```{r, echo=T}
mediation_models <- map(mediation_models_specs,
                        ~sem(., data = mediation_df)
)

map_dbl(mediation_models, AIC) %>% 
  sort()
```

# Model evaluation

* Let us now generate some data from a pure mediation model, and then fit all 8 possible versions of the mediation model to the data, and evaluate the fit.
```{r, echo=T}
mediation_df_new <- tibble(
  x = rnorm(N, sd = 2),
  m = b_m0 + b_mx * x + rnorm(N, sd = sigma_m),
  y = b_y0 + b_ym * m + rnorm(N, sd = sigma_y)
)

mediation_models_new <- map(mediation_models_specs,
                            ~sem(., data = mediation_df_new)
)

map_dbl(mediation_models_new, AIC) %>% 
  sort()
```


# Direct versus indirect effects

* In a standard linear regression model of the following kind
$$
y_i \sim N(\mu_i, \sigma^2),\quad \mu_i = \beta_0 + \beta_1 x_i,\quad \text{$i \in 1 \ldots n$},
$$
a change in any $x_{i}$ by 1 unit, i.e., $x_{i} + 1$, would always lead to a change of $\beta_1$ in the expected, i.e. the average, value of the outcome variable.
* This is easy to see. Let $x^\prime_{i} = x_{i} + 1$, and 
\begin{align*}
\mu_i = \beta_0 + \beta_1 x_{i},\quad \mu_i^\prime &= \beta_0 + \beta_1 x_{i}^\prime,\\
             &= \beta_0 + \beta_1 (x_{i} + 1),\\
             &= \beta_0 + \beta_1 x_{i} + \beta_1,\\
             &= \mu_i + \beta_1,
\end{align*}
and so $\mu^\prime - \mu = \beta_1$.
* Regardless of how many predictor variables there are in the linear regression, a change in predictor $k$ by one unit, always leads to a change $\beta_k$ in the average value of the outcome variable.

# Direct versus indirect effects


* In a mediation model, the effect of a change in the predictor $x$ on the outcome $y$ is not as simple. 
* In pure mediation model, we can write each $y_i$ and $m_i$ as follows 
\begin{align*}
y_i &= \beta_{y0} + \beta_{ym} m_i + \epsilon^y_i, \quad \epsilon^y_i \sim N(0, \sigma_y^2),\\
m_i &= \beta_{m0} + \beta_{mx} x_i + \epsilon^m_i, \quad \epsilon^m_i \sim N(0, \sigma_m^2).
\end{align*}
From this, we have 
\begin{align*}
y_i &= \beta_{y0} + \beta_{ym} \left(\beta_{m0} + \beta_{mx} x_i +  \epsilon^m_i \right) + \epsilon^y_i,\\
    &= \beta_{y0} + \beta_{ym} \beta_{m0} + \beta_{ym} \beta_{mx} x_i +  \beta_{ym} \epsilon^m_i + \epsilon^y_i,
\end{align*}
and this entails
$$
y_i \sim N(\mu_i, \beta_{ym}^2\sigma^2_m + \sigma^2_y),\quad
\mu_i = \beta_{y0} + \beta_{ym} \beta_{m0} + \beta_{ym} \beta_{mx} x_i. 
$$
* Following the same reasoning as above for the case of standard linear regression, this entails that in a pure mediation model a unit change in $x_i$ leads to a change of $\beta_{ym}\beta_{mx}$ in the expected value of $y$.

# Direct versus indirect effects

* In the case of the partial mediation model, we saw already that each $y_i$ and $m_i$ in the model can be defined as follows:
\begin{align*}
y_i &= \beta_{y0} + \beta_{ym} m_i + \beta_{yx} x_i + \epsilon^y_i,
\quad \epsilon^y_i \sim N(0, \sigma_y^2),\\
m_i &= \beta_{m0} + \beta_{mx} x_i +  \epsilon^m_i,
\quad \epsilon^m_i \sim N(0, \sigma_m^2).
\end{align*}
From this, we have 
\begin{align*}
y_i &= \beta_{y0} + \beta_{ym} (\beta_{m0} + \beta_{mx} x_i +  \epsilon^m_i) + \beta_{yx} x_i + \epsilon^y_i,\\
y_i &= \beta_{y0} + \beta_{ym}\beta_{m0} + \beta_{ym}\beta_{mx} x_i + \beta_{yx} x_i + \beta_{ym}\epsilon^m_i  + \epsilon^y_i,
\end{align*}
which entails
$$
y_i \sim N(\mu_i, \beta_{ym}^2\sigma^2_m + \sigma^2_y),\quad
\mu_i = \beta_{y0} + \beta_{ym}\beta_{m0} + \left(\beta_{ym}\beta_{mx} + \beta_{yx}\right) x_i,
$$
and following the reasoning above, this entails that unit change in $x_i$ leads to a change of $(\beta_{ym}\beta_{mx} + \beta_{yx})$ in the expected values of $y_i$.


# Direct versus indirect effects

In general in a mediation model, we have following:
$$
\underbrace{
  \overbrace{
    \beta_{ym}\beta_{mx}
    }^{\text{indirect effect}} 
    + 
  \overbrace{
    \beta_{yx}
    }^{\text{direct effect}}
  }_{\text{total effect}}.
$$
If there is no direct effect, as would be the case in pure mediation model, then the total effect is equal to the indirect effect.


# Direct versus indirect effects

* In a `lavaan` mediation model, we can create single variables that measure the direct, indirect and total effects. 
*To do so, we must first use labels for our original parameters, i.e. the coefficients, and then use the `:=` operator to create new variables that are functions of the original parameters.
```{r, echo=T}
mediation_model_spec_1 <- '
y ~ 1 + b_ym * m + b_yx * x
m ~ 1 + b_mx * x

# Define effects
indirect := b_ym * b_mx
direct   := b_yx
total    := b_yx + (b_ym * b_mx)
'
```

# Direct versus indirect effects

* We can fit this model as per usual.
```{r, echo=T}
mediation_model_1 <- sem(mediation_model_spec_1,
                         data = mediation_df)
```
In the usual parameter estimates output, we can use `dplyr::filter` to isolate these effects:
```{r, echo=T}
parameterEstimates(mediation_model_1) %>% 
  filter(label %in% c('indirect', 'direct', 'total')) %>% 
  select(label, est,pvalue:ci.upper)
```
```{r}
total_eff <- parameterEstimates(mediation_model_1) %>% 
  filter(label == 'total') %>% 
  select(est, starts_with('ci')) %>% 
  unlist() %>% 
  round(digits = 3)
```
As we can see, for example, the estimated effect for the total effect is `r total_eff['est']`, and the 95% confidence interval on this effect is $(`r total_eff['ci.lower'] `,`r total_eff['ci.upper']`)$.

# Modelling Graduate School Performance

\begin{figure}
\centering
\subfigure[]
{
\begin{tikzpicture}
\tikzstyle{every node} = [rectangle, fill=gray!30]
\tikzstyle{every path} = [-latex,thick]
\node (col) at (0, 0) {\texttt{col}};
\node (hs) at (0, 2) {\texttt{hs}};
\node (gre) at (2, 1) {\texttt{gre}};
\node (grad) at (4, 1) {\texttt{grad}};
\draw (hs) -- (gre);
\draw (col) -- (gre);
\draw (gre) -- (grad);
\draw (hs) to [bend left=30] (grad);
\draw [latex'-latex'] (hs) to (col);
\draw[-,draw opacity=0.0] (col) to [bend right=30] (grad);
\end{tikzpicture}
}
\hspace{2cm}
\subfigure[]
{
\begin{tikzpicture}
\tikzstyle{every node} = [rectangle, fill=gray!30]
\tikzstyle{every path} = [-latex,thick]
\node (col) at (0, 0) {\texttt{col}};
\node (hs) at (0, 2) {\texttt{hs}};
\node (gre) at (2, 1) {\texttt{gre}};
\node (grad) at (4, 1) {\texttt{grad}};
\draw (hs) -- (gre);
\draw (col) -- (gre);
\draw (gre) -- (grad);
\draw [latex'-latex'] (hs) to (col);
\draw (hs) to [bend left=30] (grad);
\draw (col) to [bend right=30] (grad);
\end{tikzpicture}
}
\end{figure}


# Modelling Graduate School Performance

```{r, echo=T}
grad_mediation_models_specs <- within(list(),{
  model_0 <- '
      grad ~ hs + gre
      gre ~ hs + col
  '

  model_1 <- '
      grad ~ hs + b_grad_gre*gre + b_grad_col*col
      gre ~ hs + b_gre_col*col
      
      # labels for indirect, direct, and total
      direct := b_grad_col
      indirect := b_gre_col*b_grad_gre
      total := b_grad_col + (b_gre_col*b_grad_gre)
'
})

```